Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-19 paper/paper.tex
sigplan style
fkdk

\documentclass{sig-alternate}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

@@

\subsection{Liveness Analysis for lazy languages}

@@

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn generates  a liveness  of $\{\pw,~\pw.1\sigma\}$ at  $\pi_6$. The
liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
scope of  \pu. However, for simplicity  of exposition, we  do not kill
  livenesses  of  variables  in  our formulation.   This  does  not
introduce any imprecision since \pu\  is yet to come into existence at
$\pi_6$  and  its  liveness  will  not be  consulted  by  the  garbage
collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================

@@

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path  to a  liveness environment  that is  common to  all  the program
points in the path. And a second map $\cal{P}$ maps a program point to
the set of  return paths on which the point lies.   If a program point
happens  to  fall  on more  than  one  such  path, then  the  liveness
environment of  the program  point is the  variable-wise union  of the
liveness  environments  of  the  individual  paths.   In  the  example
program, ${\cal M}(EP_1)=  \ldots$, ${\cal M}(EP_2)=\ldots$ and ${\cal
  M}(EP_3)= \ldots$.   Since ${\cal P}(pi_4)$ is  $\lbrace EP_1, EP_2,
EP_3 \rbrace$,  the liveness  of $pi_4$ is  \ldots.  In the  sequel we
shall describe how to  obtain $\cal{M}$.  The formulation of $\cal{P}$
is obvious  and we do  not describe it.   We observe that  the ${ref}$
rules neither use  ${\cal M}$ directly nor create  new labels. We only
need to consider the rules for {\LET}, {\SIF}, and {\RETURN}.

  A {\RETURN~\px}\  statement always forces an evaluation and hence a new 
  label {\em l} needs to be created and added to $\cal{M}$. A new liveness
  environment is associated with {\em l}. The demand $\sigma$ that is placed 
  on this expression is now transferred to the variable \px. {\em l} will now
  contain an entry for \px~ with liveness $\sigma$.

  An \SIF~ statement will modify  $\cal{M}$ by doing a union of 
  the maps from the {\em then} and the {\em else} part. Note that each part can contain 
  more than one label. Since the variable corresponding to the condition may not
  be part of either the {\em then} or the {\em else} part, a new label {\em l} is created 
  and added to $\cal{M}$. The only liveness associated with {\em l} is the liveness 
  of the condition variable.

  The processing rule for {\LET} is the most complex among all the rules. A {\LET} does
  not add any new label but might modify the liveness values of some variables or might add 
  liveness values for the let bound variable. Since there might be more than one {\RETURN} 
  statement following a {\LET}, there might be multiple {\em E-paths} reaching the {\LET}.
  The demand on the let variable might be different along different {\em E-paths}. Therefore,
  for each label {\em l} in $\cal{M}$, the liveness of the let-expression is calculated using 
  the liveness environment $\Lv$ corresponding to {\em l}. Since the {\LET} is lazy, 
  calculating its livness might require a fixed-point processing.

@@

First recall the classical formulation  of liveness (as sets of simple
variables)        on        three-address       instructions,        $
\mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
\mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
  strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
$\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
\in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.

Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
analogous.
% has three analogous aspects.
% \begin{itemize}
% \item
Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
returns the  incremental liveness {\em generated} by  $s$.  Because we
generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
determine which access paths of its free variables are made live.  The
cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
% The cases for
% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
$\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
argument (even if its result is never used).
This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
(note $\sigma$ may be $\{\}$).
In the opposite  sense, the  demand  of $\acar\sigma$  on
(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
that \CONS\ does not, by itself, dereference its arguments.
Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
$\mathit{ref}$
expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
Constants generate no liveness.

>>>>

\documentclass{sig-alternate}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\alan}[1]{{\color{red}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}

\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

@@

\subsection{Liveness Analysis for lazy languages}
\comment{analysis for liveness has two components - transfering
  demands on expressions to liveness of results
and propagating liveness. first discuss demand --> liveness}

%% First recall the classical formulation  of liveness (as sets of simple
%% variables)        on        three-address       instructions,        $
%% \mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
%% \mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
%%   strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
%% $\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
%% \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.

%%Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
%%analogous.
% has three analogous aspects.
% \begin{itemize}
% \item
\alan{Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
returns the  incremental liveness {\em generated} by  $s$.  Because we
generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
determine which access paths of its free variables are made live.  The
cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
% The cases for
% $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
% is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
% The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
A  demand of  $\sigma$  on  (\CAR~$x$) is  transformed  to the  demand
$\acar\sigma$  on $x$.   In  addition, \CAR\  always dereferences  its
argument  (even if  its result  is  never used).   This generates  the
liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$ (note  $\sigma$  may be
$\{\}$).   In  the opposite  sense,  the  demand  of $\acar\sigma$  on
(\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
that \CONS\ does not,  by itself, dereference its arguments.  Thirdly,
for the case  of a user-function call, a  third parameter $\Lfonly$ to
$\mathit{ref}$  expresses how  the demand  $\sigma$ on  the  result is
transformed  into demands  on its  parameters.  Constants  generate no
liveness.}

@@

Since   \pu\  is  $(\CDR~   \pw)$,  the   liveness  of   $\sigma$  for
\pu\ translates  into a demand  of $\sigma$ on $(\CDR~\pw)$,  which in
turn generates  a liveness  of $\{\pw,~\pw.1\sigma\}$ at  $\pi_6$. The
liveness of  \pu\ is  also killed at  $\pi_6$ since  it is not  in the
scope of  \pu. However, for simplicity  of exposition, we  do not kill
  livenesses  of  variables  in  our formulation.   This  does  not
introduce any imprecision since \pu\  is yet to come into existence at
$\pi_6$  and  its  liveness  will  not be  consulted  by  the  garbage
collector at this point.
%==============================================================
\begin{figure*}[t]  
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including $\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha \mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\RETURN~x),\sigma,\Lfonly) &=& \{l \mapsto x.\sigma\}, \makebox[0mm]{\hspace*{3cm} where $l$ is a new label} \\
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) &=&
        \begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \{l \mapsto  \{x.\epsilon\}\},  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}?}  & \mbox{otherwise}
                 \end{array} \\
 \makebox[0mm]{\hspace*{3.8cm} where $l$ is a new label}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \{ l \mapsto \Ldonly_{x \leftarrow s}(\Lv(x)) \mid l \mapsto \Lv
        \in \mathcal{M}\}
\mbox{ where } \mathcal{M} = \mathcal{L}(e,\sigma,\Lfonly)\\
\Ldonly_{x \leftarrow s} &=& \Lfonly_f, \mbox{where}~f~\mbox{is a new function}\\
 \makebox[0mm]{\hspace*{7cm}   
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{5.5cm} and 
     $(\ x_1\, \ldots\, x_n)$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) = 
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { d_1 \ldots d_k \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $d_1 \ldots d_k$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}
%======================================================================================

@@

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  return paths on which  the point lies.
If a  program point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(EP_1)= \ldots$, ${\cal M}(EP_2)=\ldots$ and
${\cal M}(EP_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is obvious  and we do not describe it.   We observe that the
${ref}$  rules  neither  use   ${\cal  M}$  directly  nor  create  new
labels. We  only need  to consider the  rules for {\LET},  {\SIF}, and
{\RETURN}.

  A  {\RETURN~\px}\ with a  demand $\sigma$  forces an  evaluation and
  hence a  new ${\cal M}$ needs to  be created with {\em  l} mapped to
  $\lbrace \px.\sigma \rbrace$.  An \SIF~ statement is an union of the
  ${\cal M}$-maps  from the {\em then}  and the {\em  else} part. Note
  that each part  can contain more than one  label. Since the variable
  corresponding to  the condition may not  be part of  either the {\em
    then} or the  {\em else} part, a new label {\em  l} is created and
  added to $\cal{M}$. The only liveness associated with {\em l} is the
  liveness of the condition variable. Also notice that the
  consequences of laziness---the entire statement including the
  condition is not evaluated if the demand on it is
  $\emptyset$. This results in the empty map denoted by
  $\emptyset_{\cal{M}}$.

  The processing  {\LET} rule has an elegant formulation. Recollect
  that our language permits lazy \LET s whose definitions can be
  recursive   Thus, in the expression
  $\LET~x \leftarrow~s~\IN~e$, $x$ can occur in $s$.  We model a let statement as a recursive function which captures an
  arbitrary  number
  of unrollings. We pass to this function the demand arising out of
  the liveness of $x$  at the
  beginning of $e$, and let the $the  
  that is the  most complex among  all the
  rules. A  {\LET} does  not add  any new label  but might  modify the
  liveness values of  some variables or might add  liveness values for
  the let bound variable. Since there might be more than one {\RETURN}
  statement following a {\LET},  there might be multiple {\em E-paths}
  reaching  the {\LET}.   The  demand  on the  let  variable might  be
  different along  different {\em E-paths}. Therefore,  for each label
  {\em  l}  in  $\cal{M}$,  the  liveness  of  the  let-expression  is
  calculated  using the  liveness environment  $\Lv$  corresponding to
  {\em l}.  Since the  {\LET} is lazy,  calculating its  livness might
  require a fixed-point processing.

@@

%% First recall the classical formulation  of liveness (as sets of simple
%% variables)        on        three-address       instructions,        $
%% \mathit{live}_\mathit{in}(I) = \mathit{live}_\mathit{out}(I) \setminus
%% \mathit{def\/}(I) \cup  \mathit{ref\/}(I) $,  and then note  that {\em
%%   strong liveness}  needs, when $I$ is the  instruction $z:=x+y$, that
%% $\mathit{ref}(I)$     be     refined     to    $\{x,y\}$     if     $z
%% \in\mathit{live}_\mathit{out}(I)$ and $\{\}$ otherwise.

%% Our  liveness analysis  formulated  in Figure~\ref{fig:live-judge}  is
%% analogous.
%% % has three analogous aspects.
%% % \begin{itemize}
%% % \item
%% Firstly,  the function  $\mathit{ref}$,  when given  a statement  $s$,
%% returns the  incremental liveness {\em generated} by  $s$.  Because we
%% generalize  {\em  strong}  liveness,  $\mathit{ref}$  needs  a  second
%% parameter, specifying  the demand  $\sigma$ on the  result of  $s$, to
%% determine which access paths of its free variables are made live.  The
%% cases for $(\ID\ x)$ and $(\PRIM\ x\ y)$ exemplify this.
%% % The cases for
%% % $(\CAR\ x)$, $(\CAR\ x)$ and $(\CONS\ x\ y)$ show how demand for path $\alpha \in \sigma$
%% % is transformed into a deeper or shallower path in the demand for (or liveness of) $x$ or $y$.
%% % The   operator \PRIM\ dereferences its arguments  resulting in the liveness shown.
%% A demand  of  $\sigma$  on  (\CAR~$x$)  is  transformed  to  the  demand
%% $\acar\sigma$  on  $x$.   In  addition, \CAR\  always  dereferences  its
%% argument (even if its result is never used).
%% This generates the liveness  $\{x.\epsilon\} \cup  x.\acar\sigma$
%% (note $\sigma$ may be $\{\}$).
%% In the opposite  sense, the  demand  of $\acar\sigma$  on
%% (\CONS~$x$~$y$) is  transformed to the  demand $\sigma$ on  $x$.  Note
%% that \CONS\ does not, by itself, dereference its arguments.
%% Thirdly, for the case of a user-function call, a third parameter $\Lfonly$ to 
%% $\mathit{ref}$
%% expresses how the demand $\sigma$ on the result is transformed into demands on its parameters.
%% Constants generate no liveness.
